Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
draft of related work

\steven{Talk about Conal Elliott's Automatic and Symbolic differentiation paper}

@@

While applicable to both
formal and natural languages, the existing works on categorial grammar are
frequently oriented toward giving functorial semantics to natural language in
the tradition of Montague \cite{Montague1974-MONFPS}. The Abstract Categorial
Grammar Toolkit \cite{Guillaume2024} and Grammatical Framework \cite{ranta-2011}
each mechanize aspects of Lambek calculus ; however, to the best of our knowledge,
\theoryabbv is the first mechanization designed for formal language parsing.

\steven{Contrast their interpretation of dependent types with ours. How tf do I
  say idk what they're doing. Perhaps just say that their dependent types take
  on a different interpreation and no more.}
Luo \cite{luo} describes a Lambek calculus with dependent types.

\subsubsection{Other typed approaches}
\steven{this subsection should be renamed}
The usage of a simple type system to reason about regular expressions was
introduced by Frisch and Cardelli in 2004 \cite{frischCardelli}, and later expanded
by Hengelin and Nielsen in 2011 \cite{henglein_regular_2011} to handle proofs of regular expression
containment.

\steven{Mention Neel's typed algebraic approach paper here. Say it is similar at
  first glance, as its parsing and it has types, but that he actually is representing
  a system for proving the LL(1) condition }

>>>>

@@

While applicable to both
formal and natural languages, the existing works on categorial grammar are
frequently oriented toward giving functorial semantics to natural language in
the tradition of Montague \cite{Montague1974-MONFPS}. The Abstract Categorial
Grammar Toolkit \cite{Guillaume2024} and Grammatical Framework \cite{ranta-2011}
each mechanize aspects of Lambek calculus ; however, to the best of our knowledge,
\theoryabbv is the first mechanization designed for formal language parsing first.
Moreover, we provide a novel use of inductive and dependent types inside of a
Lambek calculus.

Luo \cite{luo} describes a Lambek calculus with dependent types. However, they
do not provide an analagous denotational semantics, so it is not clear what
exactly their semantics are. Further, their contexts of non-linear variables are
ordered, while ours are not, so they provide a distinct dependent extension
whose types do not have a one-to-one correspondence with the dependent types in \theoryabbv.

\subsubsection{Typed Grammars}
The usage of a simple type system to reason about regular expressions up to weak
equivalence was
introduced by Frisch and Cardelli \cite{frischCardelli}.  Hengelin and Nielsen
\cite{henglein_regular_2011} later adapt this approach by providing a type
system for regular expressions whose semantics were in one-to-one correspondence
with the set of parse trees.
Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
semantic interpretation of grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class
of grammars.

Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
demonstrating that carries proves of the LL(1) condition for context-free
grammars. Despite being a type system over formal grammars, their interpretation
of types is very different from the semantics of \theoryabbc.
